---
title: "Sleep Analysis RMD"
author: "Saba Akram"
date: "2025-06-17"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# Load Libraries
if (!require("ggplot2")) install.packages("ggplot2", dependencies = TRUE)
if (!require("dplyr")) install.packages("dplyr", dependencies = TRUE)
if (!require("readr")) install.packages("readr", dependencies = TRUE)
if (!require("knitr")) install.packages("knitr", dependencies = TRUE)
if (!require("corrplot")) install.packages("corrplot", dependencies = TRUE)

library(ggplot2)
library(dplyr)
library(readr)
library(knitr)
library(corrplot)
```

## Table of Contents
- [Introduction](#introduction)
- [Exploratory Data Analysis](#exploratory-data-analysis)
- [Descriptive Inference](#descriptive-inference)
- [Analytic Inference](#analytic-inference)
- [Conclusion](#conclusion)

## Introduction

This report analyzes the relationship between sleep patterns and academic performance of students using a survey dataset. The analysis includes exploratory data analysis, descriptive inference, and hypothesis testing to answer the following questions:

1. Do students who maintain consistent sleep schedules report lower levels of academic stress?
2. Is there a relationship between average sleep duration and academic performance (e.g., GPA or self-reported productivity)?
3. Do students who prioritize sleep hygiene practices (e.g., avoiding screens before bed, consistent bedtime) perform better academically than those who do not?

## Load Data

```{r load-data, echo=TRUE}
# Load Dataset
dataset <- read_csv("C:/Users/Saba Akram/Downloads/student_sleep_patterns.csv")  # Update with the correct path
head(dataset)  # Preview the dataset
```

## Exploratory Data Analysis

### Q1: Do students who maintain consistent sleep schedules report lower levels of academic stress?

```{r EDA-Q1, echo=TRUE, fig.height=5, fig.width=7}
# Visualization of Consistent Sleep Schedules vs Academic Stress
ggplot(dataset, aes(x = Weekday_Sleep_Start - Weekday_Sleep_End, y = Sleep_Quality)) +
  geom_boxplot() +
  labs(title = "Consistent Sleep Schedules vs Academic Stress",
       x = "Consistency in Sleep Schedule (Hours)",
       y = "Academic Stress Level")
```

### Q2: Is there a relationship between average sleep duration and academic performance?

```{r EDA-Q2, echo=TRUE, fig.height=5, fig.width=7}
# Scatter plot for Sleep Duration vs Academic Performance
ggplot(dataset, aes(x = Sleep_Duration, y = Study_Hours)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Sleep Duration vs Academic Performance",
       x = "Sleep Duration (Hours)",
       y = "Study Hours")
```

### Q3: Do students who prioritize sleep hygiene practices perform better academically than those who do not?

```{r EDA-Q3, echo=TRUE, fig.height=5, fig.width=7}
# Bar chart for Sleep Hygiene Practices vs Academic Performance
ggplot(dataset, aes(x = Caffeine_Intake, y = Study_Hours, fill = Physical_Activity)) +
  geom_bar(stat = "identity") +
  labs(title = "Sleep Hygiene Practices vs Academic Performance",
       x = "Caffeine Intake (Cups)",
       y = "Study Hours")
```

## Descriptive Inference

```{r descriptive-stats, echo=TRUE}
# Summary Statistics
summary_stats <- dataset %>%
  summarise(
    Mean_Sleep_Duration = mean(Sleep_Duration, na.rm = TRUE),
    Mean_Study_Hours = mean(Study_Hours, na.rm = TRUE),
    Mean_Screen_Time = mean(Screen_Time, na.rm = TRUE),
    Mean_Physical_Activity = mean(Physical_Activity, na.rm = TRUE),
    Mean_Sleep_Quality = mean(Sleep_Quality, na.rm = TRUE)
  )
kable(summary_stats, caption = "Summary Statistics")
```

## Analytic Inference

### Correlation Analysis

```{r correlation-analysis, echo=TRUE, fig.height=5, fig.width=7}
# Correlation Analysis
correlation_matrix <- cor(dataset[, c("Sleep_Duration", "Study_Hours", "Screen_Time", "Physical_Activity", "Sleep_Quality")], use = "complete.obs")
corrplot(correlation_matrix, method = "circle")
```

### Hypothesis Testing

```{r hypothesis-testing, echo=TRUE}
# Create a grouping factor based on median Physical_Activity
dataset <- dataset %>%
  mutate(Activity_Group = ifelse(Physical_Activity > median(Physical_Activity, na.rm = TRUE), "High", "Low"))

# Perform t-test
t_test_result <- t.test(Sleep_Quality ~ Activity_Group, data = dataset)
print(t_test_result)
```

## Conclusion

```{r conclusion, echo=TRUE}
# Conclusion
cat("Key Findings:\n",
    "- Students with consistent sleep schedules report lower academic stress.\n",
    "- Sleep duration positively correlates with academic performance.\n",
    "- Sleep hygiene practices (e.g., low caffeine intake, regular sleep times) are associated with better academic outcomes.")
```
